Covariate analysis of MSSM reprocessed counts (with TMM normalisation)

### 2017-01-25 code/Rmd/MSSM_RNASeq_Reprocessed_TMM_BinnedCDR.Rmd
Added markdown to process reprocessed RNASeq of MSSM samples with binned CDR using TMM normalization

knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_Reprocessed_TMM.Rmd",
                                 parentId = "syn5570248",
                                 entityName = 'MSBB Reprocessed RNASeq TMM')
```

@@

library(parallel)
library(foreach)

@@

thisFileName <- 'MSSM_RNASeq_Reprocessed_TMM.Rmd'

@@

### Data preprocessing
```{r preprocess.data,cache=FALSE, echo=TRUE}
# Fix metadata
METADATA$AOD = gsub('\\+','',METADATA$AOD)
METADATA$BrodmannArea.CDR = paste(METADATA$BrodmannArea, METADATA$CDR, sep = '.')
METADATA$RIN2 = METADATA$RIN^2
levels(METADATA$batch)[levels(METADATA$batch) == ''] = 'NoBatch'

@@

Following `r length(indRemoved)` samples are removed:`r paste(indRemoved, collapse = ', ')`
```{r diagnosisPlots, fig.height=15, fig.width = 10}
# RIN
p = list()
p[[1]] = ggplot(METADATA, aes(x = BrodmannArea.CDR, y = RIN, fill = BrodmannArea)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# AOD
p[[2]] = ggplot(METADATA, aes(x = BrodmannArea.CDR, y = as.numeric(AOD), fill = BrodmannArea)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('Age At Death') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# PMI
p[[3]] = ggplot(METADATA, aes(x = BrodmannArea.CDR, y = as.numeric(PMI), fill = BrodmannArea)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# Total Reads
p[[4]] = ggplot(METADATA, aes(x = BrodmannArea.CDR, y = as.numeric(TotalReads), fill = BrodmannArea)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Total Reads') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# Mapped Reads
p[[5]] = ggplot(METADATA, aes(x = BrodmannArea.CDR, y = as.numeric(Mapped), fill = BrodmannArea)) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Mapped') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# Mapped Reads
p[[6]] = ggplot(METADATA, aes(x = BrodmannArea.CDR, y = as.numeric(rRNA.rate), fill = BrodmannArea)) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('rRNA Rate') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

@@

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrodmannArea.CDR")
ContCovariates <- c("RIN", "RIN2", "AOD","PMI", "TotalReads", "Mapped", "rRNA.rate")

@@

Remove genes that have less than 1 cpm counts in at least 50% of samples per BrodmannArea.CDR
```{r cpmnormalisation}
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('sampleIdentifier') %>%
  dlply(.(BrodmannArea.CDR), .fun = function(mtd, count){
    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$sampleIdentifier],
                                                     MIN_GENE_CPM=1, 
                                                     MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
    processed.counts$filteredExprMatrix$genes
    }, COUNT)
genesToAnalyze = unlist(genesToAnalyze) %>% unique()

@@

```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = summary(COVARIATES$BrodmannArea.CDR) %>%
  rownameToFirstColumn('BrodmannArea.CDR') %>%
  tidyr::separate(BrodmannArea.CDR, c('BrodmannArea', 'CDR'), sep = 5) %>%
  tidyr::spread(CDR, DF)
kable(tmp)
```

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=factor(CDR), color=as.factor(BrodmannArea), size=as.numeric(RIN)))
p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:7)
# p <- p + geom_text(aes(label= sampleIdentifier), size=4, hjust=0)
pl[[1]] <- p

p <- ggplot(plotdata, aes(x=PC2, y=PC3))
p <- p + geom_point(aes(shape=factor(CDR), color=as.factor(BrodmannArea), size=as.numeric(RIN)))
p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:7)
# p <- p + geom_text(aes(label= sampleIdentifier), size=4, hjust=0)
pl[[2]] <- p

@@

COVARIATES.tmp = COVARIATES[, c('batch','RACE','SEX','BrodmannArea.CDR')]
COVARIATES.tmp = sapply(COVARIATES.tmp, as.numeric)
col.mat = labels2colors(cbind(tree,COVARIATES.tmp))
col.mat = gsub('\\.1', '', col.mat)

@@

# Find significant covariates
adjustCovars = preAdjustedSigCovars$significantCovars
```
Significant covariates to adjust at FDR 0.1 are `r paste(gsub('_','\\\\_',adjustCovars), collapse= ',')`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=25, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
```{r iterative.adjusted.voom.normalisation, results='asis'}
postAdjustCovars = c('BrodmannArea.CDR', 'batch', 'SEX')
randomCovariates = 'individualIdentifier'

@@

tmp <- paste('Using following covariates in the final model:', modelStr, 'and Donor_ID as random effects')
```
`r tmp`
### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
Adding "BrodmannArea.CDR" back to the residuals
```{r addVarsBack}
varsToAddBackIn <-  grep("BrainRegion.Diagnosis", colnames(FIT$coefficients), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(FIT$coefficients[,varsToAddBackIn]))
```

@@

pl = list()
p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=factor(CDR), color=as.factor(BrodmannArea), size=as.numeric(RIN)))
p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:7)
# p <- p + geom_text(aes(label= sampleIdentifier), size=4, hjust=0)
pl[[1]] <- p

p <- ggplot(plotdata, aes(x=PC2, y=PC3))
p <- p + geom_point(aes(shape=factor(CDR), color=as.factor(BrodmannArea), size=as.numeric(RIN)))
p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:7)
# p <- p + geom_text(aes(label= sampleIdentifier), size=4, hjust=0)
pl[[2]] <- p

@@

# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host = "www.ensembl.org")

@@

VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('gene.id') %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=20}
# Fit contrast
cntr = c()
for (BA in unique(METADATA$BrodmannArea)){
  tmp = combn(paste0('BrodmannArea.CDR', BA, '.', c(0,0.5,1,2,3,4,5)),2)
  cntr = c(cntr, apply(tmp, 2, paste, collapse = '-'))
}

@@

# Get differnetial expression
DE = lapply(1:84, function(i, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION){
  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
}, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION)
names(DE) = gsub('BrodmannArea.CDR','',colnames(contrast))

DE1 = plyr::ldply(DE, .id = 'Comparison') %>%
  dplyr::mutate(Comparison1 = Comparison) %>%
  tidyr::separate(Comparison1, c('from.state','to.state'), sep = '-') %>%
  tidyr::separate(from.state, c('BrodmannArea','From.CDR'), sep = 5) %>%
  tidyr::separate(to.state, c('BrodmannArea','To.CDR'), sep = 5) %>%
  tidyr::unite(Name, From.CDR, To.CDR, sep = '-')

p <- ggplot(DE1, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
p <- p + geom_hline(yintercept = -log10(0.05), color = 'red')
p <- p + geom_vline(xintercept = log2(1.2), color = 'red')
p <- p + geom_vline(xintercept = -log2(1.2), color = 'red')
p <- p + facet_grid(BrodmannArea~.+Name)
p

tmp = DE1 %>% 
  dplyr::filter(adj.P.Val <= 0.05) %>%
  group_by(BrodmannArea, Name) %>%
  summarise(count = length(unique(genes))) %>%
  tidyr::spread(BrodmannArea, count)
kable(tmp)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'MSBB Reprocessed RNASeq TMM', parentId = parentId)
annotations(CODE) = list(
  dataType = 'mRNA',
  dataSubType = 'geneExp',
  summaryLevel = 'gene',
  assay	 = 'Reprocessed RNAseq',

>>>>

knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_Reprocessed_TMM_BinnedCDR.Rmd",
                                 parentId = "syn5570248",
                                 entityName = 'MSBB Reprocessed RNASeq TMM Binned CDR')
```

@@

library(parallel)
library(foreach)
library(doMC)
library(foreach)
library(doParallel)

@@

thisFileName <- 'MSSM_RNASeq_Reprocessed_TMM_BinnedCDR.Rmd'

@@

### Data preprocessing
```{r preprocess.data,cache=FALSE, echo=TRUE}
# Fix metadata
METADATA$AOD = gsub('\\+','',METADATA$AOD)
METADATA$Dx[METADATA$CDR <= 1] = 'ND'
METADATA$Dx[METADATA$CDR == 2 | METADATA$CDR == 3] = 'MD'
METADATA$Dx[METADATA$CDR >= 4] = 'SD'
METADATA$BrodmannArea.Dx = paste(METADATA$BrodmannArea, METADATA$Dx, sep = '.')
METADATA$RIN2 = METADATA$RIN^2
levels(METADATA$batch)[levels(METADATA$batch) == ''] = 'NoBatch'

@@

Following `r length(indRemoved)` samples are removed:`r paste(indRemoved, collapse = ', ')`
```{r diagnosisPlots, fig.height=15, fig.width = 10}
# RIN
p = list()
p[[1]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = RIN, fill = BrodmannArea)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# AOD
p[[2]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(AOD), fill = BrodmannArea)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('Age At Death') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# PMI
p[[3]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(PMI), fill = BrodmannArea)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# Total Reads
p[[4]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(TotalReads), fill = BrodmannArea)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Total Reads') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# Mapped Reads
p[[5]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(Mapped), fill = BrodmannArea)) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Mapped') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# Mapped Reads
p[[6]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(rRNA.rate), fill = BrodmannArea)) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('rRNA Rate') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

@@

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrodmannArea.Dx")
ContCovariates <- c("RIN", "RIN2", "AOD","PMI", "TotalReads", "Mapped", "rRNA.rate")

@@

Remove genes that have less than 1 cpm counts in at least 50% of samples per BrodmannArea.Dx
```{r cpmnormalisation}
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('sampleIdentifier') %>%
  dlply(.(BrodmannArea.Dx), .fun = function(mtd, count){
    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$sampleIdentifier],
                                                     MIN_GENE_CPM=1, 
                                                     MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
    processed.counts$filteredExprMatrix$genes
    }, COUNT)
genesToAnalyze = unlist(genesToAnalyze) %>% unique()

@@

```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = summary(COVARIATES$BrodmannArea.Dx) %>%
  rownameToFirstColumn('BrodmannArea.Dx') %>%
  tidyr::separate(BrodmannArea.Dx, c('BrodmannArea', 'Dx'), sep = 5) %>%
  tidyr::spread(Dx, DF)
kable(tmp)
```

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=factor(Dx), color=as.factor(BrodmannArea), size=as.numeric(RIN)))
p <- p + theme_bw() + theme(legend.position="left") + scale_shape_manual(values = 1:7)
# p <- p + geom_text(aes(label= sampleIdentifier), size=4, hjust=0)
pl[[1]] <- p

p <- ggplot(plotdata, aes(x=PC2, y=PC3))
p <- p + geom_point(aes(shape=factor(Dx), color=as.factor(BrodmannArea), size=as.numeric(RIN)))
p <- p + theme_bw() + theme(legend.position="left") + scale_shape_manual(values = 1:7)
# p <- p + geom_text(aes(label= sampleIdentifier), size=4, hjust=0)
pl[[2]] <- p

@@

COVARIATES.tmp = COVARIATES[, c('batch','RACE','SEX','BrodmannArea.Dx')]
COVARIATES.tmp = sapply(COVARIATES.tmp, as.numeric)
col.mat = labels2colors(cbind(tree,COVARIATES.tmp))
col.mat = gsub('\\.1', '', col.mat)

@@

# Find significant covariates
adjustCovars = preAdjustedSigCovars$significantCovars
```
Significant covariates to adjust at FDR 0.1 are `r paste(gsub('_','\\\\_',adjustCovars), collapse= ',')`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=25, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
```{r iterative.adjusted.voom.normalisation, results='asis'}
postAdjustCovars = c('BrodmannArea.Dx', 'batch', 'SEX')
randomCovariates = 'individualIdentifier'

@@

tmp <- paste('Using following covariates in the final model:', modelStr, 'and Donor_ID as random effects')
```
`r tmp`
### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
Adding "BrodmannArea.Dx" back to the residuals
```{r addVarsBack}
varsToAddBackIn <-  grep("BrodmannArea.Dx", colnames(FIT$coefficients), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(FIT$coefficients[,varsToAddBackIn]))
```

@@

pl = list()
p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=factor(Dx), color=as.factor(BrodmannArea), size=as.numeric(RIN)))
p <- p + theme_bw() + theme(legend.position="left") +scale_shape_manual(values = 1:7)
# p <- p + geom_text(aes(label= sampleIdentifier), size=4, hjust=0)
pl[[1]] <- p

p <- ggplot(plotdata, aes(x=PC2, y=PC3))
p <- p + geom_point(aes(shape=factor(Dx), color=as.factor(BrodmannArea), size=as.numeric(RIN)))
p <- p + theme_bw() + theme(legend.position="left") +scale_shape_manual(values = 1:7)
# p <- p + geom_text(aes(label= sampleIdentifier), size=4, hjust=0)
pl[[2]] <- p

@@

# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host = "dec2016.archive.ensembl.org")

@@

VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('gene.id') %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=20}
# Fit contrast
cntr = c()
for (BA in unique(METADATA$BrodmannArea)){
  tmp = combn(paste0('BrodmannArea.Dx', BA, '.', c('ND','MD','SD')),2)
  cntr = c(cntr, apply(tmp, 2, paste, collapse = '-'))
}

@@

# Get differnetial expression
DE = lapply(1:12, function(i, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION){
  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1], confint = T)
}, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION)
names(DE) = gsub('BrodmannArea.Dx','',colnames(contrast))

DE1 = plyr::ldply(DE, .id = 'Comparison') %>%
  dplyr::mutate(Comparison1 = Comparison) %>%
  tidyr::separate(Comparison1, c('from.state','to.state'), sep = '-') %>%
  tidyr::separate(from.state, c('BrodmannArea','From.Dx'), sep = 5) %>%
  tidyr::separate(to.state, c('BrodmannArea','To.Dx'), sep = 5) %>%
  tidyr::unite(Dx, From.Dx, To.Dx, sep = '-')

p <- ggplot(DE1, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
p <- p + geom_hline(yintercept = -log10(0.05), color = 'red')
p <- p + geom_vline(xintercept = log2(1.2), color = 'red')
p <- p + geom_vline(xintercept = -log2(1.2), color = 'red')
p <- p + facet_grid(BrodmannArea~.+Dx)
p

tmp = DE1 %>% 
  dplyr::filter(adj.P.Val <= 0.05) %>%
  group_by(BrodmannArea, Dx) %>%
  summarise(count = length(unique(genes))) %>%
  tidyr::spread(BrodmannArea, count)
kable(tmp)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'MSBB Reprocessed RNASeq TMM Binned CDR', parentId = parentId)
annotations(CODE) = list(
  dataType = 'mRNA',
  dataSubType = 'geneExp',
  summaryLevel = 'gene',
  assay	 = 'Reprocessed RNAseq',
